---
title: "ST558 Homework 5"
format: html
editor: visual
---

## Task 1
```{r}
## Join via code provided in zip
d1=read.table("student-mat.csv",sep=";",header=TRUE)
d2=read.table("student-por.csv",sep=";",header=TRUE)

d3=merge(d1,d2,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet"))
print(nrow(d3))
```

```{r}
library(tidyverse)
math_data <- read_delim("student-mat.csv", delim = ";", col_names = TRUE)
por_data <- read_delim("student-por.csv", delim = ";", col_names = TRUE)
inner_join(math_data, por_data)  # does not print any rows
```

```{r}
## This will be used in further steps
combined_data <- inner_join(math_data, por_data, 
                            by = c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason", "guardian", "traveltime", "studytime", "failures", "schoolsup", "famsup", "activities", "nursery", "higher", "internet", "romantic", "famrel", "freetime", "goout", "Dalc", "Walc", "health"))
```

```{r}
## Changing variables of interest into factors
math_data <- math_data |>
  mutate(across(c("Pstatus", "Mjob", "Fjob", "internet"), as.factor))

por_data <- por_data |>
  mutate(across(c("Pstatus", "Mjob", "Fjob", "internet"), as.factor))

combined_data <- combined_data |>
  mutate(across(c("Pstatus", "Mjob", "Fjob", "internet"), as.factor))
```

## Task 2
```{r}
## one-way contingency table
table(combined_data$internet, useNA = "always")
```
The number 48 indicates that 48 students do not have access to internet at home
```{r}
## two-way contingency table
table(combined_data$Pstatus, combined_data$Mjob, useNA = "always")
```
The number 2 indicates that 2 students have parents who live apart where the mother works in healthcare
```{r}
## three-way contingency table
table(combined_data$Fjob, combined_data$Mjob, combined_data$internet, useNA = "always")
```
The number 2 in the second row of the first table means that of students with no access to internet at home, 2 of them have a father who works in healthcare while also having a mother that works as a teacher

```{r}
## conditional two-way table using filter()
table_no_internet <- combined_data |>
  filter(internet == "no")

table(table_no_internet$Pstatus, table_no_internet$Mjob, useNA = "always")
```

```{r}
## sub-setting three-way contingency table to print two-way table
three_way_table <- table(combined_data$Pstatus, combined_data$Mjob, combined_data$internet, useNA = "always")
three_way_table[ , , "no"]
```

```{r}
## two-way table using dplyr
combined_data |>
  group_by(Pstatus, Mjob) |>
  summarize(count = n()) |>
  pivot_wider(names_from = Mjob, values_from = count)
```

```{r}
## stacked bar graph
ggplot(data = combined_data, aes(x = internet, fill = Mjob)) +
  geom_bar() +
  labs(x = "Internet Accessability", y = "Count of mothers with each job", title = "Student's Mothers' Jobs based on Internet Accessability") +
  scale_fill_discrete("Mother's Job")
```

```{r}
## side-by-side bar graph
ggplot(data = combined_data, aes(x = internet, fill = Mjob)) +
  geom_bar(position = "dodge") +
  labs(x = "Internet Accessability", y = "Count of mothers with each job", title = "Student's Mothers' Jobs based on Internet Accessability") +
  scale_fill_discrete("Mother's Job")
```

```{r}
## finding center & spread of numeric variables
combined_data |>
  summarize(across(c("G3.x", "G3.y", "absences.x"),
                   list("mean" = ~ mean(.x, na.rm = TRUE), "sd" = ~ sd(.x, na.rm = TRUE)),
                   .names = "{.fn}_{.col}"))
```

```{r}
## finding center & spread of numeric variables while sub-setting for students with internet at home
combined_data |>
  filter(internet == "yes") |>
  summarize(across(c("G3.x", "G3.y", "absences.x"),
                   list("mean" = ~ mean(.x, na.rm = TRUE), "sd" = ~ sd(.x, na.rm = TRUE)),
                   .names = "{.fn}_{.col}"))
```

```{r}
## finding center & spread across a single grouping variable
combined_data |>
  group_by(internet) |>
  summarize(across(c("G3.x", "G3.y", "absences.x"),
                   list("mean" = ~ mean(.x, na.rm = TRUE), "sd" = ~ sd(.x, na.rm = TRUE)),
                   .names = "{.fn}_{.col}"))
```

```{r}
## finding center & spread across two grouping variables
combined_data |>
  group_by(Pstatus, Mjob) |>
  summarize(across(c("G3.x", "G3.y", "absences.x"),
                   list("mean" = ~ mean(.x, na.rm = TRUE), "sd" = ~ sd(.x, na.rm = TRUE)),
                   .names = "{.fn}_{.col}"))
```

```{r}
## correlation matrix of all numeric variables
combined_data |>
  select(age, absences.x, absences.y, G1.x, G1.y, G2.x, G2.y, G3.x, G3.y) |>
  cor()
```

```{r}
## histogram
ggplot(data = combined_data, aes(x = G3.x, fill = internet)) +
  geom_histogram() +
  labs(x = "Final Math Grade", y = "Number of Students", title = "Distribution of Student Final Math Grades")

ggplot(data = combined_data, aes(x = G3.y, fill = internet)) +
  geom_histogram() +
  labs(x = "Final Portuguese Grade", y = "Number of Students", title = "Distribution of Student Final Portuguese Grades")
```

```{r}
## kernel density
ggplot(data = combined_data, aes(x = G3.x, fill = internet)) +
  geom_density() +
  labs(x = "Final Math Grade", y = "Number of Students", title = "Density of Student Final Math Grades")

ggplot(data = combined_data, aes(x = G3.y, fill = internet)) +
  geom_density() +
  labs(x = "Final Portuguese Grade", y = "Number of Students", title = "Density of Student Final Portuguese Grades")
```

```{r}
## boxplot
ggplot(data = combined_data, aes(x = G3.x, fill = internet)) +
  geom_boxplot() +
  labs(x = "Final Math Grade", y = "Number of Students", title = "Boxplot of Student Final Math Grades")

ggplot(data = combined_data, aes(x = G3.y, fill = internet)) +
  geom_boxplot() +
  labs(x = "Final Portuguese Grade", y = "Number of Students", title = "Boxplot of Student Final Portuguese Grades")
```

```{r}
## scatterplot
ggplot(data = combined_data, aes(x = absences.x, y = G3.x, color = internet)) +
  geom_point() +
  labs(x = "Math Class Absenses", y = "Final Math Grade", title = "Number of Math Class Absenses vs Final Math Grade") +
  geom_jitter(width = 0.2, alpha = 0.3)

ggplot(data = combined_data, aes(x = G1.x, y = G3.x, color = internet)) +
  geom_point() +
  labs(x = "First Period Math Grade", y = "Final Math Grade", title = "First Period Math Grade vs Final Math Grade") +
  geom_jitter(width = 0.2, alpha = 0.3)
```

```{r}
## scatterplot with faceting
ggplot(data = combined_data, aes(x = absences.x, y = G3.x, color = internet)) +
  geom_point() +
  labs(x = "Math Class Absenses", y = "Final Math Grade", title = "Number of Math Class Absenses vs Final Math Grade") +
  geom_jitter(width = 0.2, alpha = 0.3) +
  facet_wrap(~ Pstatus)

ggplot(data = combined_data, aes(x = G1.x, y = G3.x, color = internet)) +
  geom_point() +
  labs(x = "First Period Math Grade", y = "Final Math Grade", title = "First Period Math Grade vs Final Math Grade") +
  geom_jitter(width = 0.2, alpha = 0.3) +
  facet_wrap(~ Pstatus)
```

```{r}
## scatterplot with faceting with 2 categorical variables
ggplot(data = combined_data, aes(x = absences.x, y = G3.x, color = internet)) +
  geom_point() +
  labs(x = "Math Class Absenses", y = "Final Math Grade", title = "Number of Math Class Absenses vs Final Math Grade") +
  geom_jitter(width = 0.2, alpha = 0.3) +
  facet_grid(Mjob ~ Pstatus)

ggplot(data = combined_data, aes(x = G1.x, y = G3.x, color = internet)) +
  geom_point() +
  labs(x = "First Period Math Grade", y = "Final Math Grade", title = "First Period Math Grade vs Final Math Grade") +
  geom_jitter(width = 0.2, alpha = 0.3) +
  facet_wrap(Mjob ~ Pstatus)
```